---
title: "Graphing with R and ggplot"
author:  "Jim Tyson"
date: "`r Sys.Date()'"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# Introduction

We will be using `gapminder` data.  You will need some tidyverse functions, including ggplot.  

```{r}
# libraries

library(ggplot2)
library(gapminder)
library(tidyverse)


# data

gm <- gapminder

gm <- gm %>% 
  mutate(european = continent == "Europe")

```

# [The R graph gallery](https://r-graph-gallery.com/index.html)

Base R plots work (usually) on vectors of data.  So for example we can create to vectors 

```{r}

xdata = c(1:100)

ydata = xdata^2

```
and make a scatter plot of xdata against ydata:

```{r}

plot(xdata,ydata, type = "l", main="Basic XY Scatter Plot",xlab="The X axis vector",ylab="The Y axis vector")

```

There are other base graph types, but they all take a 'canvas' approach    

# In the ggplot world, plots are made of **layers**.

```{r}

gm %>% 
  ggplot(aes(x=lifeExp))+
  geom_histogram()

```

A plot must have a layer with data, a mapping of data series to plot dimensions and position data (often default positionining).  For visualising the data a geometric object must be selected from ggplot2s collection of `geom_`s.  Each `geom_` is associated with a default `statistic`,  conversely each `statistic` has a default geometric representation.

In this first graph, the data is in the object `gm`, and the variable `lifeExp` is mapped to the x axis.  This is drawn as a histogram.  The `geom_histogram` uses the default statistic *frequency*.

##   Data

Ggplot expects data in a dataframe (or some analog, such as a `tibble`.)  You can supply the dataframe to the initial `ggplot` call, or in a subsequent layer.  So, you can plot data as points in one layer and then in another layer plot different data as a line.

## Aesthetic mapping

The `aes()` specification of the ggplot call, specifies how data elements are matched to visual properties of `geom_`s.  For example, that some data series is mapped to the y axis, or some measure is represented by the size of data points.  You can use the `aes()` options to add colour coding to data using `fill` or `colour`.  `aes()` properties set in the `ggplot()` call will cascade to further layers unless overridden.

```{r}

gm %>% 
  select(gdpPercap,continent,year) %>% 
  group_by(continent) %>% 
  filter(year==2007) %>% 
  ggplot()+
  geom_boxplot(aes(x=continent,y=gdpPercap))

gm %>% 
  ggplot(aes(x=year,y=lifeExp,fill=continent))+
  geom_smooth(method="lm")+
  facet_grid(rows =vars(continent))

````

##  Geometric object

There are several geometries that can be used in plots, including columns, points, lines and boxes.  You can see all the `geoms` at https://ggplot2.tidyverse.org/reference/.


-   Statistical transformations

As examples, the histogram has frequency as its default statistic.  If you do not specify any statistical transformation the default is used 

## Scales

Control the properties of the data axes - including labels and limits.

## Position adjustments

Position adjustments are made to resolve overlapping of `geom_`s.

##  Faceting

A graph that displays multiple data series can be broken into a grid of seperate sub-plots by `facet_grid` 

!("C:/Users/ccaajim/OneDrive - University College London/Pictures/rgraphing/ggplot2_blocks.png")
x   
# a layer consists of data, stat, geom and position

# a layer must have data, which can be a dataframe/tibble, unlike

# base R plots which work with vectors, each 'geom' has a default 'stat' and

# vice versa: geom_bar -\> count; geom_point -\> identity

# reference: <https://ggplot2.tidyverse.org/reference/#stats>

# <https://tinyurl.com/ggplotstatref>

# \~/Pictures/ggplotstatref-400(1).png

# mostly I recommend you use ggplot2 for graphing but I will show

# a few base R graphs as examples

# some base R graphs

# dot plot

plot(gm\$gdpPercap)

# add a line and change colour

plot(gm$gdpPercap,type="o", col="blue") lines(gm$lifeExp\*1000,
type="o") \# add title title(main="GDP Per Capita Population and life
expectancy e10\^3", col.main="red", font.main=4)

# selecting data and building a basic graph using ggplot

# and tidyverse functions

gm %\>% select(gdpPercap,continent,year) %\>% group_by(continent) %\>%
filter(year==2007) %\>% ggplot()+ geom_boxplot(aes(x=continent,
y=gdpPercap))

# some gapminder data graphed

# <http://euclid.psych.yorku.ca/www/psy6135/tutorials/gapminder.html>

# we begin a scatter graph definition and store it in 'p'

# we specify a data frame and a mapping for x and y

# axes

# we add the geometry for the graph with +

# how might we improve this? Could some transformation help?

p \<- ggplot(gm, aes(x = gdpPercap, y = lifeExp)) + geom_point() p \<-
p + scale_x\_log10() p \<- p + geom_point(aes(color=continent)) \# title

# too much! redundant color specification doesn't help!

# a plot should always have proper labels

p + geom_point(aes(col=continent, shape=continent, size=lifeExp)) p \<-
p + geom_point(aes(col=continent), size=3) p \<- p+ggtitle("Life
Expectancy against GDP Per Capita Population") p \<- p+xlab("GDP Per
Capita in USD") p \<- p + theme_bw() p p+theme_classic()

# can try to fit a line to this data

# the default method for smoothing is lowess and it includes shading for

# SE

p + geom_smooth() \# change the method and remove the SE, reset the line
width and color... p + geom_point() + geom_smooth(lwd=2, se=FALSE,
method="lm", col="red") \# add color back on the geom_point aes p +
geom_point(aes(color = continent)) + geom_smooth() \# add the continent
coloring to the whole graph - it's inherited \# by all following geoms!
p + aes(color = continent) + geom_point() + geom_smooth() p + aes(color
= continent) + geom_point() + geom_smooth(se=F, lwd=2)

# box plots

# some limitations of point plotting...

p \<- ggplot(gm, aes(continent, lifeExp)) p+geom_point() \# overplotting
\# trying transparency and jitter... p + geom_point(alpha=1/4) p +
geom_jitter() p + geom_jitter(alpha=1/2)

p + geom_boxplot() \# we could try using both... p + geom_jitter() +
geom_boxplot() p + geom_boxplot(outlier.colour = "red") +
geom_jitter(alpha=1/2)

# boxplot with a categorical variable

# and an alternative, the violin plot

p \<- ggplot(gm, aes(x=continent, y=gdpPercap)) p + geom_boxplot() p +
geom_violin(aes(fill=continent))

# extreme values are a problem here

# try with lifeExp

p \<- ggplot(gm, aes(x=european, y=lifeExp)) p + geom_boxplot() p +
geom_violin()

# Graphing types

## Density graphs

### box plot

# to help build plots incrementally, we will save each layer and modification in

# 'p'

p \<- ggplot(gm, aes(european, lifeExp, fill = continent))

p + geom_boxplot()

# an alternative to the box plot is a violin plot

p + geom_violin()

### histogram

p \<- ggplot(gm, (aes(x=pop))) p + geom_histogram()

# notes on graphs

# <https://stackoverflow.com/questions/10197738/add-a-footnote-citation-outside-of-plot-area-in-r>
